Reasoning Graphically about Quantum Computation

### 2008-02-24 papers/reasoning-graphically/paper.tex
added make file generator to convert EPS to PDF files
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1356 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

% Boldified names of useful categories
%
\newcommand{\vectfd}[1]{% category of finite dimensional vector spaces over some field
\ensuremath{\textbf{Vect}^{\mathrm{fd}}_{#1}}\xspace}
\newcommand{\vectfdc}{% category of finite dimensional vector spaces over complexes
\vectfd{\mathbb{C}}}
\newcommand{\catRel}{% the category of sets and relations
\ensuremath{\textbf{Rel}}\xspace}
\newcommand{\catSet}{% the category of sets and functions
\ensuremath{\textbf{Set}}\xspace}
\newcommand{\catCat}{% the category of (small) categories
\ensuremath{\textbf{Cat}}\xspace}
\newcommand{\catInvCat}{% the category of involutive categories
\ensuremath{\textbf{InvCat}}\xspace}
\newcommand{\catComCl}{% the category of compact closed  categories
\ensuremath{\textbf{ComCl}}\xspace}
\newcommand{\catSComCl}{% the category of strongly compact closed  categories
\ensuremath{\textbf{SComCl}}\xspace}
\newcommand{\catGrph}{% the category of graphs
\ensuremath{\textbf{Grph}}\xspace}
\newcommand{\qubit}{% the category of qubits
\ensuremath{\textbf{Qubit}}\xspace}
\newcommand{\fdhilb}{% the category of finite dimensional hilbert space
\ensuremath{\textbf{FDHilb}}\xspace}
\newcommand{\catInvCom}{% the category of involutive categories
\ensuremath{\textbf{InvCom}}\xspace}
\newcommand{\catCom}{% the category of compact closed  categories
\ensuremath{\textbf{Com}}\xspace}
%%%% End of Ross's lovely categorcal macros--------
%%--------------------------------------------

@@

Unfortunately. or fortunately, few examples of interesting
compact closed categories are freely generated.  In practice we work
with cateories presented as generators and \emph{equations}.  For
representing quantum computations, Coecke and Duncan
\cite{Coecke:2008jo} propose the following collection of generators:
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] $\epsilon_Z : Q \to I$, $\delta_Z : Q \to Q \otimes Q$,
  $\alpha_Z : Q \to Q$, $\epsilon_X : Q \to I$, $\delta_X : Q \to Q \otimes Q$,
  $\alpha_X : Q \to Q$ and $H:Q\to Q$.
\end{description}
subject to  the following equations:
\begin{gather}
  (\delta \otimes \id{Q}) \circ \delta 
   =   ( \id{Q} \otimes \delta) \circ \delta  
  \tag{A}\label{comonoid-assoc}
  \\
  (\epsilon \otimes \id{Q}) \circ \delta 
  = (\id{Q} \otimes \epsilon)\circ \delta 
  = \id{Q} \tag{U} \label{comonoid-unit}
  \\
  \delta = \sigma \circ \delta \tag{C}\label{comonoid-comm}
  \\
  \delta^\dag \circ \delta = \id{Q} \tag{I}\label{isometry}
  \\
  (\delta^\dag \otimes \id{Q}) \circ (\id{Q}\otimes \delta)
  = \delta \circ \delta^\dag \tag{F}\label{frobenius}
\end{gather}

>>>>

% Boldified names of useful categories
%
\newcommand{\vectfd}[1]{% category of finite dimensional vector spaces over some field
\ensuremath{\textbf{Vect}^{\mathrm{fd}}_{#1}}\xspace}
\newcommand{\vectfdc}{% category of finite dimensional vector spaces over complexes
\vectfd{\mathbb{C}}}
\newcommand{\catRel}{% the category of sets and relations
\ensuremath{\textbf{Rel}}\xspace}
\newcommand{\catSet}{% the category of sets and functions
\ensuremath{\textbf{Set}}\xspace}
\newcommand{\catCat}{% the category of (small) categories
\ensuremath{\textbf{Cat}}\xspace}
\newcommand{\catInvCat}{% the category of involutive categories
\ensuremath{\textbf{InvCat}}\xspace}
\newcommand{\catComCl}{% the category of compact closed  categories
\ensuremath{\textbf{ComCl}}\xspace}
\newcommand{\catSComCl}{% the category of strongly compact closed  categories
\ensuremath{\textbf{SComCl}}\xspace}
\newcommand{\catGrph}{% the category of graphs
\ensuremath{\textbf{Grph}}\xspace}
\newcommand{\qubit}{% the category of qubits
\ensuremath{\textbf{Qubit}}\xspace}
\newcommand{\fdhilb}{% the category of finite dimensional hilbert space
\ensuremath{\textbf{FDHilb}}\xspace}
\newcommand{\catInvCom}{% the category of involutive categories
\ensuremath{\textbf{InvCom}}\xspace}
\newcommand{\catCom}{% the category of compact closed  categories
\ensuremath{\textbf{Com}}\xspace}
%%%% End of Ross's lovely categorcal macros--------

%%%%  use this to include graphics in the right place.
\newcommand{\inlinegraphic}[2]{
  %% todo -- make this thing calculate the height 
  %% itself based on a global scaling factor
  \dimendef\grafheight=255\dimendef\grafvshift=254
  \grafheight=#1
  \grafvshift=-0.5\grafheight
  \advance\grafvshift by 0.5ex
  \raisebox{\grafvshift}{\includegraphics[height=\grafheight]{#2}\xspace}
}

%%--------------------------------------------

@@

Few examples of interesting compact closed categories are freely
generated.  In practice we work with categories presented as
generators and \emph{equations}.  For 
representing quantum computations, Coecke and Duncan
\cite{Coecke:2008jo} propose the following collection of generators:
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
We represent these two families, and their adjoints, by colours in our
graphical notation:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{alpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
%  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
%  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inline
Each family $(\delta, \epsilon, \alpha)$ is governed by a set of
equations defining a forms a \emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima} with a
\emph{phase group}.  The requisite equations state that $(\delta,
\epsilon )$ form a cocommutative comonoid:
\begin{gather}
  (\delta \otimes \id{Q}) \circ \delta 
   =   ( \id{Q} \otimes \delta) \circ \delta  
  \tag{A}\label{comonoid-assoc}
  \\
  (\epsilon \otimes \id{Q}) \circ \delta 
  = (\id{Q} \otimes \epsilon)\circ \delta 
  = \id{Q} \tag{U} \label{comonoid-unit}
  \\
  \delta = \sigma \circ \delta. \tag{C}\label{comonoid-comm}
\end{gather}
Their adjoints $(\delta^\dag,\epsilon^\dag)$ form a commutative
monoid':
\begin{gather}
  \delta^\dag \circ (\delta^\dag \otimes \id{Q})
  =    \delta^\dag   \circ ( \id{Q} \otimes \delta^\dag) 
  \tag{A}\label{monoid-assoc}
  \\
  \delta^\dag  \circ (\epsilon^\dag \otimes \id{Q}) 
  = \delta^\dag \circ (\id{Q} \otimes \epsilon^\dag)
  = \id{Q} \tag{U} \label{monoid-unit}
  \\
  \delta^\dag = \delta^\dag \circ
  \sigma. \tag{C}\label{monoid-comm}.
\end{gather}
Futher, these jointly satisfy the \emph{isometry} and \emph{frobenius} laws:
\begin{gather}
  \delta^\dag \circ \delta = \id{Q} \tag{I}\label{isometry}
  \\
  (\delta^\dag \otimes \id{Q}) \circ (\id{Q}\otimes \delta)
  = \delta \circ \delta^\dag. \tag{F}\label{frobenius}
\end{gather}
